---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

## Loading and preprocessing the data

```{r load-and-preprocess}
unzip("activity.zip")
rawData <- read.csv("activity.csv")
summary(rawData)
```

## What is mean total number of steps taken per day?

```{r mean-total-steps-per-day}
totalStepsPerDay <- rawData %>% group_by(date) %>%
  summarise(totalStepsPerDay=sum(steps, na.rm=TRUE), .groups="drop") %>%
  pull(totalStepsPerDay)
hist(totalStepsPerDay,
     xlab="Total steps per day",
     main="Frequency of total steps per day",
     breaks=20)
mean(totalStepsPerDay)
median(totalStepsPerDay)
```

## What is the average daily activity pattern?

```{r average-daily-activity}
dailyMeanSteps <- rawData %>% group_by(interval) %>%
  summarise(dailyMean=mean(steps, na.rm=TRUE), .groups="drop")
plot(
  dailyMeanSteps,
  type="l",
  xlab="5-minute interval in HHMM format",
  ylab="Average number of steps across all days")
```

The 5-minute interval with the maximum average number of steps across all
the days in the dataset is 08:35-08:40 with 206 steps on average.

```{r max-interval}
dailyMeanSteps[which.max(dailyMeanSteps$dailyMean), ]
```

## Imputing missing values

As in the above summary, the number of NAs is:
```{r count-of-missing-values}
length(rawData[is.na(rawData$steps), 1])
```

I am going to replace the missing values by `mean` over the same hour. If there
are no values at all to compute the mean, I will put 0 as the number of steps.

```{r imputting-missing-values}
data <- data.frame(rawData)
data$datetime <- as.POSIXct( # add standard datetime column
  paste(
    data$date,
    paste(
      as.character(data$interval %/% 100),
      as.character(data$interval %% 100),
      sep=":")))
data$hour <- format(data$datetime, "%Y-%m-%d %H") # add "mod hour" column
data <- merge( # add meanSteps column (over the same hour)
  data,
  data %>% group_by(hour=format(datetime, "%Y-%m-%d %H")) %>%
    summarise(meanSteps=mean(steps,na.rm=TRUE), .groups="drop"),
  by="hour")
# replace NA steps by meanSteps
data[is.na(data$steps), "steps"] <- data[is.na(data$steps), "meanSteps"]
# replace NaN steps (due to 0 non-NA steps within the same hour) by 0
data[is.nan(data$steps), "steps"] <- 0
data <- subset(data, select=-c(hour, meanSteps))
summary(data)
```

### The mean total number of steps taken per day after imputting missing values

```{r mean-total-steps-per-day-2}
totalStepsPerDay2 <- data %>% group_by(date) %>%
  summarise(totalStepsPerDay2=sum(steps), .groups="drop") %>%
  pull(totalStepsPerDay2)
hist(totalStepsPerDay2,
     xlab="Total steps per day",
     main="Frequency of total steps per day",
     breaks=20)
mean(totalStepsPerDay2)
median(totalStepsPerDay2)
```

In the end, given the strategy of replacing the NAs by means, there is no
significant impact on the total number of steps per day, their `mean` and
`median`.

## Are there differences in activity patterns between weekdays and weekends?

```{r weekdays, fig.width=8, fig.height=8}
data$dayFactor <- factor(
  weekdays(data$datetime, abbreviate=TRUE),
  levels=c("Mon", "Tue", "Wed", "Thu", "Fri", "Sam", "Sun"),
  labels=c("Weekday", "Weekday", "Weekday", "Weekday", "Weekday",
           "Weekend", "Weekend"))
par(mfrow=c(2, 1))
par(mar=c(4, 4, 0, 0))
plot(data %>% filter(dayFactor == "Weekday") %>%
       group_by(interval) %>%
       summarise(m=mean(steps), .groups="drop"),
     type="l",
     xlab="5-minute interval in HHMM format",
     ylab="Weekday avg")
plot(data %>% filter(dayFactor == "Weekend") %>%
       group_by(interval) %>%
       summarise(m=mean(steps), .groups="drop"),
     type="l",
     xlab="5-minute interval in HHMM format",
     ylab="Weekend avg")
```